|

### 2023-05-03 manuscript/manuscript_1.0.Rmd
New ways to asses the effect of frequency dependent selection on the signal

mtext(text = "Trait value",cex = cexAxis,line = 1,side = 2,las=0)
legend("topright",legend = c(expression(alpha[s])),col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanAlpha_"),runChoi)],
         col=colGenesLin,lty = 2,type="l",lwd=5)

@@

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c(expression(beta[s])),cex=cexAxis,
       col=colGenesLin[2],lwd=2,bty = "n")
legend("topleft",legend = c("A.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanBeta_"),runChoi)],
         col=colGenesLin[2],lty = 2,type="l",lwd=4)

@@

# meansClustmp<-tempPop[,.(alphaMean=mean(alpha),betaMean=mean(beta)),by=orderClus]
# reacNormpClust<-sapply(meansClustmp[,orderClus], function(x){
#   logist(rangx,alpha = meansClustmp[orderClus==x,alphaMean],
#          beta = meansClustmp[orderClus==x,betaMean])})
# matplot(x = rangx,y = reacNormpClust,col=colRuns[meansClustmp$orderClus],
#          type='l',cex.axis=cexAxis,
#          xlab="",ylab="",ylim=c(0,1.1),lty=1,
#          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  reacNormInd<-sapply(tempPop[indId %in%randInds,indId], function(x){
    logist(rangx,alpha = tempPop[indId==x,alpha],
           beta = tempPop[indId==x,beta])})
  matplot(x = rangx,y = reacNormInd,
          type='l',cex.axis=cexAxis,
          xlab="",ylab="",ylim=c(0,1.1),lty=1,col=1,
          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  mtext(text = seqYlabDown[count],cex = cexAxis,line = 1.5,side = 2,las=0)
  mtext(text = seqXlabDown[count],cex = cexAxis,line = 1.3,side = 1,las=0)
  legend("topleft",legend =  c(paste0("B.",count)),bty = "n",inset = -0.06)
  }

@@

The amount of information that agents are able to collect through learning 
throughout their life time can strongly change the outcome of 
evolutionary dynamics. In the simulations presented in figure 
\ref{fig:handicap} and \ref{fig:noBadge}, 
individuals learned with high speed of learning ($\mathrm{A} = 0.4$) and interacted 
repeatedly along their lifetime (on average 2000 interactions). When we reduced 
the number of interactions that individuals have over their lifetime to 300 
on average, we saw a drastic increase in the phenotypic and genetic
variation present in the population. Fig. \ref{fig:branching} shows 
changes in the distribution of the sender reaction norm parameters
along evolutionary time, assuming the badge is cost-free. 
In panels A1 and 2, darker areas show trait values with high frequency in the population. 
Populations start monomorphic with a value of zero on both the intercept and
the slope, and mutations quickly build up a normal distribution around the starting 
value. Within the first 2000 generations, the unimodal distribution in 
the intercept ($\alpha$) splits into a bimodal one. 
Later in evolutionary time one of the peaks splits further, 
so at the end of the evolutionary simulation the distribution of the intercept 
($\alpha$) in the population shows three peaks. In the case of $\beta$, the 
peaks in the distribution are not so clear-cut, but it is clear that there is an 
increase in the variance of the distribution. These changes in the 
distribution of the parameters of the sender reaction norm 
imply that individuals can generally be classified into three 
distinct types (\ref{fig:branching}). Two types express a flat reaction 
norm with extreme values for the badge size, meaning that their badge size 
is not informative of their quality, whereas the third type shows 
intermediate badge sizes determined by the quality of the individual 
(Fig. \ref{fig:branching} B.3-4). 
In this particular simulation, the values of the slope in the informative group
are negative, which implies that badge size correlates negatively with quality. 
Furthermore, the receiver reaction norms developed through learning, particularly those 
of individuals of intermediate quality, respond to the signal of their 
sender type by increasing the probability of retreating from a fight with 
individuals with smaller badges (Fig. \ref{fig:branching}).

@@

mtext(text = "Trait value",cex = cexAxis,line = 1,side = 2,las=0)
legend("topright",legend = c(expression(alpha[s])),col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("A"),cex=cexAxis,bty = "n",inset = -0.06)

@@

  points(data=popOneInd[(get(shortSce)==PAr&time>max(time)*0.8)&
                          seed == repsScen[count]],
       alpha~beta,col=multDiscrPallet[seed+1],cex=2,pch=20)

@@

@@

mtext(text = "Trait value",cex = cexAxis,line = 1,side = 2,las=0)
legend("topright",legend = c(expression(alpha[s])),col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c(expression(beta[s])),cex=cexAxis,
       col=colGenesLin[2],lwd=2,bty = "n")
legend("topleft",legend = c("B"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

@@

```{r learnEvol,echo=FALSE,out.width='100%',fig.height=6,fig.align='center',fig.show='hold',fig.cap= "The evolution of costly signals. Portrait of the evolutionary dynamics of the sender code with snapsots of both sender and receiver codes just as in fig. \\ref{fig:handicap}. The middle panels show changes in the distribution of values for $\\alpha$ and $\\beta$ along evolutionary time. Panels above and below correspond to snapshots of the sender and receiver codes, respectively, generation time of the snapshots are indicated by the grey lines in the middle panels. Similarly to the evolutionary dynamics in Fig. \\ref{fig:handicap}, first the badge size evolves towards its minimum value. Then, the slope evolves positive values determining an increasing reaction norm. In contrast to \\ref{fig:handicap}, the normal distribution splits up into two modes. This translates into two types of reaction norms assorting in the population where both code for a positive relation between quality and badge size."}

@@

mtext(text = "Trait value",cex = cexAxis,line = 1.8,side = 2,las=0)
legend("topright",legend =
      expression(paste("intercept (",alpha[s],")")),col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = expression(paste("Slope (",beta[s],")")),cex=cexAxis,
       col=colGenesLin[2],lwd=2,bty = "n")
legend("topleft",legend = c("A.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanBeta_"),runChoi)],
         col=colGenesLin[2],lty = 1,type="l",lwd=4)

@@

mtext(text = "Trait value",cex = cexAxis,line = 1.8,side = 2,las=0)
legend("topright",legend = "Speed of learning",col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("B.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

evolDist(indData = pop[seed==runChoi],variable = "initAct",nbins = 10,
         pal = pal_dist, nlevels=10,cexAxis = 1.5,xlab="",ylab = "",
         xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,initAct])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c("Behavioural tendency before learning"),cex=cexAxis,
       col=colGenesLin[2],lwd=2,bty = "n")
legend("topleft",legend = c("B.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,initAct]))),
                      lty=1,col = "grey",lwd=2)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanInitAct_"),runChoi)],
         col=colGenesLin[2],lty = 1,type="l",lwd=4)

@@

# meansClustmp<-tempPop[,.(alphaMean=mean(alpha),betaMean=mean(beta)),by=orderClus]
# reacNormpClust<-sapply(meansClustmp[,orderClus], function(x){
#   logist(rangx,alpha = meansClustmp[orderClus==x,alphaMean],
#          beta = meansClustmp[orderClus==x,betaMean])})
# matplot(x = rangx,y = reacNormpClust,col=colRuns[meansClustmp$orderClus],
#          type='l',cex.axis=cexAxis,
#          xlab="",ylab="",ylim=c(0,1.1),lty=1,
#          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  reacNormInd<-sapply(tempPop[indId %in%randInds,indId], function(x){
    logist(rangx,alpha = tempPop[indId==x,alpha],
           beta = tempPop[indId==x,beta])})
  matplot(x = rangx,y = reacNormInd,
          type='l',cex.axis=cexAxis,
          xlab="",ylab="",ylim=c(0,1.1),lty=1,col=1,
          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  mtext(text = seqYlabDown[count],cex = cexAxis,line = 1.9,side = 2,las=0)
  mtext(text = seqXlabDown[count],cex = cexAxis,line = 1.7,side = 1,las=0)
  legend("topleft",legend = c(paste0("C.",count)),bty = "n",inset = -0.06)
  }

>>>>

mtext(text = "Trait value",cex = cexAxis,line = 1,side = 2,las=0)
legend("topright",legend = c(expression(alpha[s])), bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

@@

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c(expression(beta[s])),cex=cexAxis,bty = "n")
legend("topleft",legend = c("A.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanBeta_"),runChoi)],
         col=colGenesLin,lty = 1,type="l",lwd=4)

@@

# meansClustmp<-tempPop[,.(alphaMean=mean(alpha),betaMean=mean(beta)),by=orderClus]
# reacNormpClust<-sapply(meansClustmp[,orderClus], function(x){
#   logist(rangx,alpha = meansClustmp[orderClus==x,alphaMean],
#          beta = meansClustmp[orderClus==x,betaMean])})
# matplot(x = rangx,y = reacNormpClust,col=colRuns[meansClustmp$orderClus],
#          type='l',cex.axis=cexAxis,
#          xlab="",ylab="",ylim=c(0,1.1),lty=1,
#          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  reacNormInd<-sapply(tempPop[indId %in%randInds,indId], function(x){
    logist(rangx,alpha = tempPop[indId==x,alpha],
           beta = tempPop[indId==x,beta])})
  matplot(x = rangx,y = reacNormInd,
          type='l',cex.axis=cexAxis,
          xlab="",ylab="",ylim=c(0,1.1),lty=1,col="grey",
          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  mtext(text = seqYlabDown[count],cex = cexAxis,line = 1.5,side = 2,las=0)
  mtext(text = seqXlabDown[count],cex = cexAxis,line = 1.3,side = 1,las=0)
  legend("topleft",legend =  c(paste0("B.",count)),bty = "n",inset = -0.06)
  }

@@

The amount of information that agents are able to collect through learning 
throughout their life time can strongly change the outcome of 
evolutionary dynamics. In the simulations presented in figure 
\ref{fig:handicap} and \ref{fig:noBadge}, 
individuals learned with high speed of learning ($\mathrm{A} = 0.4$) and interacted 
repeatedly along their lifetime (on average 2000 interactions). When we reduced 
the number of interactions that individuals have over their lifetime to 300 
on average, we saw a drastic increase in the phenotypic and genetic
variation present in the population. Fig. \ref{fig:branching} shows 
changes in the distribution of the sender reaction norm parameters
along evolutionary time, assuming the badge is cost-free. 
In panels A1 and 2, darker areas show trait values with high frequency in the population. 
Populations start monomorphic with a value of zero on both the intercept and
the slope, and mutations quickly build up a normal distribution around the starting 
value. Within the first 2000 generations, the unimodal distribution in 
the intercept ($\alpha$) splits into a bimodal one. 
Later in evolutionary time one of the peaks splits further, 
so at the end of the evolutionary simulation the distribution of the intercept 
($\alpha$) in the population shows three peaks. In the case of $\beta$, the 
peaks in the distribution are not so clear-cut, but it is clear that there is an 
increase in the variance of the distribution. These changes in the 
distribution of the parameters of the sender reaction norm 
imply that individuals can generally be classified into three 
distinct types (Fig. \ref{fig:branching}). Two types express a flat reaction 
norm with extreme values for the badge size, meaning that their badge size 
is not informative of their quality, whereas the third type shows 
intermediate badge sizes determined by the quality of the individual 
(Fig. \ref{fig:branching} B.3-4). 
In this particular simulation, the values of the slope in the informative group
are negative, which implies that badge size correlates negatively with quality. 
Furthermore, the receiver reaction norms developed through learning, particularly those 
of individuals of intermediate quality, respond to the signal of their 
sender type by increasing the probability of retreating from a fight with 
individuals with smaller badges (Fig. \ref{fig:branching}).

@@

mtext(text = "Trait value",cex = cexAxis,line = 1,side = 2,las=0)
legend("topright",legend = c(expression(alpha[s])),
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("A"),cex=cexAxis,bty = "n",inset = -0.06)

@@

  points(data=popOneInd[(get(shortSce)==PAr&time>max(time)*0.8)&
                          seed == repsScen[count]],
       alpha~beta,col=multDiscrPallet[seed+1],cex=1,pch=20)

@@

The action of natural selection and genetic drift do not override the conditions 
for the evolutionary branching described earlier. Given how crucial naive 
behaviour is for the diversification process described earlier, we tested whether 
evolutionary processes could lead key parameters of the learning process toward
values that prevented the genetic variation to build up. 
Figure \ref{fig:learnEvol} shows the dynamics from a set of simulations where we 
allow for the coevolution of the reaction norm as well as key parameters of
the learning module, namely the speed of learning ($A$) and the behavioural 
tendency expressed by individuals before learning. Top panels in 
figure \ref{fig:learnEvol} show that there is diversification process in this 
coevolutionary scenario as well. That is because, as is shown in the middle 
panels (B) of the same figure, the learning parameters stay within the range 
necessary for the diversification process. The speed of learning 
(Fig. \ref{fig:learnEvol} B.1) stays well above zero throughout the simulation. 
In contrast, the initial behavioral tendency maintain values around zero. 
Thus, evolutionary processes acting on the learning parameteres 
did not override the conditions necessary for frequency dependent to boost 
genetic variation.

@@

mtext(text = "Trait value",cex = cexAxis,line = 1,side = 2,las=0)
legend("topright",legend = c(expression(alpha[s])),bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c(expression(beta[s])),cex=cexAxis,bty = "n")
legend("topleft",legend = c("A.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

@@

```{r learnEvol,echo=FALSE,out.width='100%',fig.height=6,fig.align='center',fig.show='hold',fig.cap= "Coevolutionary dynamics of both the signal reaction norm and key parameters of the learning module. Dynamics are portrait as changes in the distribution of values assorting in the population. In the top panels (A)  evolutionary dynamics of the sender code; A.1 shows the intercept and A.2 the slope of the logistic reaction norm. Both of these parameters experience evolutionary branching, whereby in the end three distincs types are assorting in the population. The middle panels (B) show changes in the distribution of values for the speed of learning ($A$;B1) and the behavioural tendency before learning ($y_z^0$; B2) along evolutionary time. The speed of learning does not change drastically, it maintains values well above 0 (black line). In contrast, the initial behavioural tendency does not evolve away from 0 (black line). Both of these conditions favour evolutionary branching in the signal reaction norm. Bottom Panels correspond to snapshots of the sender code. Generation time of the snapshots are indicated by the grey lines in the top and middle panels."}

@@

mtext(text = "Trait value",cex = cexAxis,line = 1.8,side = 2,las=0)
legend("topright",legend =
      expression(paste("intercept (",alpha[s],")")),bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = expression(paste("Slope (",beta[s],")")),cex=cexAxis,
       bty = "n")
legend("topleft",legend = c("A.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanBeta_"),runChoi)],
         col=colGenesLin,lty = 1,type="l",lwd=4)

@@

mtext(text = "Trait value",cex = cexAxis,line = 1.8,side = 2,las=0)
legend("topright",legend = "Speed of learning", bty = "n",cex=cexAxis)
legend("topleft",legend = c("B.1"),cex=cexAxis,bty = "n",inset = -0.06)

@@

evolDist(indData = pop[seed==runChoi],variable = "initAct",nbins = 10,
         pal = pal_dist, nlevels=10,cexAxis = 1.5,xlab="",ylab = "",
         xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,initAct])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c("Naive behavioural tendency"),cex=cexAxis,
       bty = "n")
legend("topleft",legend = c("B.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,initAct]))),
                      lty=1,col = "grey",lwd=2)

@@

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanInitAct_"),runChoi)],
         col=colGenesLin,lty = 1,type="l",lwd=4)

@@

# meansClustmp<-tempPop[,.(alphaMean=mean(alpha),betaMean=mean(beta)),by=orderClus]
# reacNormpClust<-sapply(meansClustmp[,orderClus], function(x){
#   logist(rangx,alpha = meansClustmp[orderClus==x,alphaMean],
#          beta = meansClustmp[orderClus==x,betaMean])})
# matplot(x = rangx,y = reacNormpClust,col=colRuns[meansClustmp$orderClus],
#          type='l',cex.axis=cexAxis,
#          xlab="",ylab="",ylim=c(0,1.1),lty=1,
#          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  reacNormInd<-sapply(tempPop[indId %in%randInds,indId], function(x){
    logist(rangx,alpha = tempPop[indId==x,alpha],
           beta = tempPop[indId==x,beta])})
  matplot(x = rangx,y = reacNormInd,
          type='l',cex.axis=cexAxis,
          xlab="",ylab="",ylim=c(0,1.1),lty=1,col="grey",
          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  mtext(text = seqYlabDown[count],cex = cexAxis,line = 1.9,side = 2,las=0)
  mtext(text = seqXlabDown[count],cex = cexAxis,line = 1.7,side = 1,las=0)
  legend("topleft",legend = c(paste0("C.",count)),bty = "n",inset = -0.06)
  }
